---
title: "Data Visualisation with ggplot2 - Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question
Analyse data for Hilton International Hotels to provide strategy recommendations that will lead to revenue growth.

### b) Defining the Metric for Success
The solutions to the following questions will help us answer our research question:

* When is the best time of year to book a hotel room?
* When is the optimal length of stay in order to get the best daily rate?
* How will you know if a hotel was likely to receive a disproportionately high number of special requests?

### c) Understanding the context
You are a Data Scientist working for Hilton International Hotels and need to help the management decide on
the strategies that will lead to revenue growth.

You have a dataset containing information on the various chain of hotels, including when customers made
the booking, length of stay, the number of adults, children, or babies, and the number of available parking spaces, among other things.

### d) Recording the Experimental Design
We will first perform data exploration, data cleaning, and analysis to come with appropriate
recommendations.

### e) Data Relevance
The data provided was relevant to the research question since it answers the research questions


## 2. Reading the Data
```{r}
library(tidyverse)
# install.packages("reshape")
library(reshape)
```

```{r}
# Load the data below
# --- 
# Dataset url = https://bit.ly/2WvQbhJ
# --- 
# YOUR CODE GOES BELOW
# 
hotel_booking_df <- read.csv("hotel_bookings.csv", header = TRUE, na.strings=c("NULL",NA))

```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
head(hotel_booking_df, 5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(hotel_booking_df, 5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(hotel_booking_df, 10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(hotel_booking_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 
glimpse(hotel_booking_df)
```

```{r}

```

Record your general observations below:

Observation 1:
There are about 120 thousand records in the data set with 32 fields for every record. Each field seem to have the correct data type
Observation 2: 
Some of the Values in the dataset are either null or missing


## 3. External Data Source Validation
The data is originally from the article Hotel Booking Demand Datasets, by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume 22, February 2019.


## 4. Data Preparation
### Performing Data Cleaning
```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
str(hotel_booking_df)
sum(is.na(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1: Correct datatypes are assigned for the different fields
Observation 2: There are 129425 missing values in our data set

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 
# hotel_booking_df[duplicated(hotel_booking_df), ]
sum(duplicated(hotel_booking_df))
```

We choose to keep the duplicates because we don't have a unique identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 
which(colSums(is.na(hotel_booking_df))  == ncol(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1: There are no columns in which all the data is null

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW

which(rowSums(is.na(hotel_booking_df)) == nrow(hotel_booking_df))
```

We observe the following from our dataset:

Observation 1:  there is no row in which all the data is null

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 

head(hotel_booking_df)

# Select the desired columns
hotel_booking_df2 <- hotel_booking_df %>%
  na.omit() %>%
  select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, adults, children, babies, is_repeated_guest, previous_cancellations, previous_bookings_not_canceled, days_in_waiting_list, total_of_special_requests)
hotel_booking_df2

# Create a correlation matrix
corr_matrix <- cor(hotel_booking_df2, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
corr_df

# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "black") +
  geom_text(
    aes(label = round(value, 2)),
    color = "pink"
  ) +
  coord_fixed() +
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    limit = c(-1,1)
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

There is weak correlation (both positive and negative) between the various fields with the exception of "previous_bookings_not_canceled" and "is_repeated_guest" where we have a strong positive correlation 


```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 
hotel_booking_df = select(hotel_booking_df, -company)
head(hotel_booking_df)
```

From the data variable description we see that the Distribution Channel category that tells us about Booking distribution.

The term “TA” means “Travel Agents”
The term “TO” means “Tour Operators”
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

hotel_booking_df$agent[is.na(hotel_booking_df$agent)] <- "TO"
hotel_booking_df
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

hotel_booking_df <- subset(hotel_booking_df, adults!=0 | children!=0 | babies!=0)
hotel_booking_df
```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#
hotel_booking_df$children[is.na(hotel_booking_df$children)] <- round(mean(hotel_booking_df$children, na.rm = TRUE), 0)
hotel_booking_df
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
sum(is.na(hotel_booking_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 
# hotel_booking_df$children <- lapply(hotel_booking_df$children, as.numeric)
as.integer(hotel_booking_df$children)
hotel_booking_df
```


## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot

hotel_booking_df %>%
  group_by(reservation_status) %>%
  summarise(is_canceled = sum(is_canceled))

ggplot(hotel_booking_df, aes(x= reservation_status, y= is_canceled))+
  geom_col(stat = "identity", width =  0.5, fill= "blue") +
  labs(
    title = "Reservation Status",
    x= "Status",
    y= "Number"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Ariel Black", color="purple")
  )
  
```


```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

resort_city_hotel <- hotel_booking_df %>%
  count(hotel)
resort_city_hotel

ggplot(data=resort_city_hotel, aes(x="", y=n, fill=hotel)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") +
  geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette="Set2")+
  labs(
    fill = "Hotel",
    title = "Booking Between Resort Hotel and City Hotel"
  )+
  theme_void()
```


```{r}
# 3. What was the percentage of booking for each year?
# ---
# 

year_booking <- hotel_booking_df %>%
  count(arrival_date_year)
year_booking

ggplot(data=year_booking, aes(x="", y=n, fill=factor(arrival_date_year))) +
  geom_bar(
    stat="identity", width=0.5, color = "cyan"
  ) +
  coord_polar("y") +
    geom_text(
    aes(label = paste0(round(n / sum(n) * 100, 2), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +
  labs(
    fill = "Years",
    title = "Booking for each year") +
  theme_void()
```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 
busy_months <- hotel_booking_df %>%
  count(arrival_date_month)
busy_months

ggplot(busy_months, aes(x= arrival_date_month, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "magenta") +
  labs(
    title = "Bookings per Month",
    x= "Month",
    y= "Number of bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )
```


```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#

hotel_booking_df

countries <- hotel_booking_df %>%
  group_by(country) %>%
  summarise(guests = sum(adults)+sum(children)+sum(babies)) %>%
  # arrange(desc(guests)) %>%
  top_n(3)
countries

ggplot(countries, aes(x= country, y= guests))+
  geom_col(stat = "identity", width =  0.8, fill= "blue") +
  labs(
    title = "Top 3 Countries per Guests",
    x= "Countries",
    y= "Number of Guests"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="black")
  )+
  coord_flip()
```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 

stay_length <- hotel_booking_df %>%
  group_by(hotel) %>%
  summarise(total_stay_days = mean(stays_in_weekend_nights)+mean(stays_in_week_nights))
stay_length

ggplot(data = stay_length, aes(x= hotel, y= total_stay_days))+
  geom_col(stat = "identity", width =  0.8, fill= "dark grey") +
  labs(
    title = "Days Stayed per Hotel",
    x= "Hotel Type",
    y= "Number of Days Stayed"
  ) +
  theme(
    plot.title = element_text(hjust = 0, size=16, family="Arial", color="blue")
  )+
  theme_classic()
```


```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#
accommodation <- hotel_booking_df %>%
  count(reserved_room_type)
accommodation

ggplot(data = accommodation, aes(x= reserved_room_type, y= n))+
  geom_col(stat = "identity", width =  0.8, fill= "brown") +
  labs(
    title = "Bookings per Accomodation Type",
    x= "Room Type",
    y= "Number of Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="blue")
  )+
coord_flip()
```

### 5.b) Recommendations
From the above analysis, below are our recommendations:
A lot of cancellation do happen accross the hotels while only a few are a no show. Reasons for cancellation should be investigated so as to minimize on lost revenue

Twice as much guests prefer city hotels  to resort hotels. Management should invest in city hotels to ensure they have sufficnent capacity 

Booking are not consistent year on year and management should be prepared for flactuations in number of guests each year

August is the busisest month and management must be prepared to handle the high number of guests in busy months to tap in revenue
Incentives can be provided during the less busy months to encourage bookings 

Most guests come from Portugal, Great Britain and France



## 6. Challenging your Solution
In this step, we review our solution and implement approaches that could potentially provide a better outcome. In our case, we could propose the following question that wasn't answered in our solution because it couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 
not_busy_months <- hotel_booking_df %>%
  count(arrival_date_month)%>%
  top_n(-3)
not_busy_months

ggplot(not_busy_months, aes(x= arrival_date_month, y= n))+
  geom_col(stat = "identity", width =  0.5, fill= "#3B7A57") +
  labs(
    title = "Least Bookings per Month",
    x= "Month",
    y= "Number of bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Arial", color="#4B3621")
  )
```

Our observations:
November, December, and January are the least busy months. This means most of the facilities in the hotel remain underutilized and management can seek to provide offers that incentivize guests to make bookings during these months 

How does this observation tie to our solution?
This observation helps management with an opportunity to further grow their revenue by attracting guests even during the low seasons

## 7. Follow up questions
During this step, you rethink and propose other ways that you can improve your solution.

a). Did we have the right data?
Yes, we had the right data which was from a credible source
b). Do we need other data to answer our question?
provided data was sufficient to answer our research questions, however, other data that could provide more insights such as reasons for the high number of cancellations would have been of great help
c). Did we have the right question?
Research questions formulated were adequate to provide sound recommendations


